Word Orders Trade Off Predictability and Memory Load Across Languages

### 2019-02-17 paper.tex
updating
paper

\title{Word Orders Trade Off Predictability and Memory Load Across Languages}%  Word Order Minimizes  Across Languages
\author{Michael Hahn}
\date{2018}

@@

\section{Two Memory-Surprisal Tradeoffs}

With minimal assumptions, we will use information theory to derive two tradeoffs:

(1) a tradeoff between \emph{speaker memory} and (listener) surprisal

(2) a tradeoff between \emph{listener memory} and (listener) surprisal

In the second part of the paper, we examine whether word orders in natural language optimise these two tradeoffs.

@@

What can we say about the amount of information that the speaker has to store when producing utterances -- that is, how many bits does it take to encode the speaker's  memory states $S_t$ and $L_t$?
Our first theoretical result sheds light on this; we will show that the amount of bits needed to encode the memory state is related to the decay of mutual information over distances.

@@

Using this concept, we can now state our first main result, concerning the memory load of a speaker: %, and of a listener performing optimal incremental prediction:

\begin{proposition}\label{prop:lower-bound}
	A speaker producing samples from $(X_t)_{t \in \mathbb{Z}}$ needs to maintain at least
\begin{equation}\label{eq:memory-bound}
M := \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
\end{equation}
bits of memory on average.

\end{proposition}

Also can relax this to imprecise production: A speaker with low memory must deviate strongly from the language, measured in KL divergence.

@@

\subsection{Languages can trade off Memory Load and Predictability}
%\subsection{Speakers can can minimize Memory Load at Fixed Information Rate}

We have derived a general lower bound on the memory load of a speaker in Proposition~\ref{prop:lower-bound}.
We now relate the speaker's memory to the surprisal experienced by the listener.

The average surprisal of a language $(X_t)_{t \in \mathbb{Z}}$, as experienced by a listener performing optimal prediction, is equal to the entropy rate $H[X_t|X_{<t}]$.
This can be written as the unigram entropy minus the mutual information between a single observation $X_t$ and the past:
\begin{equation}
	H[X_t|X_{<t}] = H[X_t] - I[X_t, X_{<t}]
\end{equation}
where the unigram entropy $H[X_t]$ only depends on the frequencies of individual words, independently of the language's word order.

We will refer to the mutual information $I[X_t, X_{<t}]$ as the \emph{predictability} of the process.
The predictability $P := I[X_t, X_{<t}]$ is a lower bound on $M = I[X_{\geq t}, X_{<t}]$, and thus on speaker memory:
\begin{equation}
P \leq M
\end{equation}
and thus\footnote{More generally 
$H[X_t|X_{<t}] \geq \frac{1}{T} \left(H[X_{1\dots T}] - M\right)$ for any $T$. Thus, the following relation holds for any $T$: $$T \cdot \text{ListenerSurprisal} \geq H[X_{1\dots T}] - \text{SpeakerMemory}$$}
\begin{equation}
	Listener\ Surprisal \geq	H[X_t|X_{<t}] \geq H[X_t] - M \geq H[X_t] - Speaker\ Memory
\end{equation}
This means that, when we consider different languages that share a single unigram distribution, memory and surprisal trade off:
Languages that are highly predictable  require a good amount of speaker memory; a language that can be produced with very small memory cannot be very predictable.
In particular, a language that requires zero speaker memory must have zero predictability and maximal average surprisal.

We thus obtain a tradeoff between two different types of costs facing speakers and listeners: The \emph{speaker's} memory trades off with the \emph{listeners'} surprisal.

We can visualize this by placing languages into a \emph{memory-surprisal-plane} (Figure~\ref{fig:toy-mis}). Each language is assigned its entropy rate as its $y$ coordinate, and the memory load necessary to achieve optimal surprisal as its $x$ coordinate.
We will continue to assume that all languages share a single unigram distribution -- this will in particular apply when studying different word order variations of the same language.

The plane can be separated into an \emph{achievable} region, populated by languages, and an \emph{unachievable} region in which no languages can occur.
No language can have zero surprisal with zero memory -- the corner of the plane thus must fall in the unachievable region.
The boundary between the two regions is the \emph{Pareto frontier}.
Languages that fall on the boundary optimally trade off surprisal and memory; they cannot further decrease one without increasing the other.
The smallest possible values for this boundary consist of the points satisfying $P=M$, indicated by a straight diagonal line in Figure~\ref{fig:toy-mis}.
Only part of the achievable region will be populated by languages that could plausibly be natural languages: For instance, we may expect natural languages to have long-term dependencies that will prevent them from being at the absolute Pareto frontier.
We hypothesize that, if languages optimize the tradeoff between memory and surprisal, they will be close to the optimal boundary of a region populated by a `comparison class' of languages that share essential characteristics of natural language, but have other word order rules.
%When We will specify a class of baseline languages that we will argue to form a reasonable comparison class for natural language.
%Within this, we hypothesize that natural language is close to the Pareto frontier.

\begin{figure}
\includegraphics[width=0.45\textwidth]{toy/pareto-scheme.pdf}
%
	\caption{Memory-Surprisal Tradeoff. The straight line represents the Pareto frontier for unconstrained languages. Languages resembling natural languages inhabit a subset of the achievable region (grey oval). Its lower-left boundary represents the Pareto frontier relevant for natural language. We hypothesize that natural language lies close to this frontier.   }\label{fig:toy-mis}
\end{figure}

\subsection{Listeners can trade off Memory Load and Surprisal}\label{sec:listener}

%TODO redo curves so that x-axis is surprisal, y-axis the necessary memory

We have analyzed the memory load of a speaker and related it to the listener's surprisal.
We now analyze memory from the perspective of the listener, who needs to maintain information about the past to predict the future.
Similar to the listener maintains a memory state $L_t$; again, there are no assumptions about the memory architecture and the nature of its computations.
%We assume that the listener has no access to the speaker's state beyond what the speaker has already uttered.
We only make two basic assumptions about the flow of information:

@@

\subsection{Relation to Theories of Sentence Processing}

@@

\paragraph{Statistical Complexity}
There are deep connections between our formalization of speaker and listener memory and studies of dynamic systems in the Physics literature.
Our formalization of listener memory, under the assumption of optimal prediction, is equivalent to the notion of \emph{Statistical Complexity} \citep{crutchfield-inferring-1989}.
Speaker memory corresponds to \emph{Generative Complexity} \cite{loehr-non-sufficient-2008, loehr-predictive-2010}.
The tradeoff between listener memory and surprisal is formally equivalent to the \emph{Recursive Information Bottleneck} considered by \cite{still-information-2014}.
The quantity in (\ref{eq:memory-bound}) is equal to the \emph{excess entropy}, and it has long been known to bound statistical and generative complexity \citep{crutchfield-inferring-1989}.
However, the link between memory and locality provided by our series expansion in (\ref{eq:memory-bound}) appears to be a novel contribution.

@@

\section{Heavy NP Shift}

Phenomenon linked to locality/memory (Hawkins)

Is the listener's memory-surprisal tradeoff optimized by Heavy NP Shift?

Penn Treebank

Trained 20 language models counterfactual version

then estimate memory on real and reverse orderings

3) Describe the key dependent variable(s) specifying how they will be measured.
Collect all instances of a VP consisting of a verb, and (in some order) an NP and a PP, from the Penn Treebank (WSJ) training set. Calculate the listener's memory-surprisal tradeoff on the words belonging to the NP and the PP.

5) Specify exactly which analyses you will conduct to examine the main question/hypothesis.
(1) Is the listener's memory-surprisal tradeoff significantly better on real as opposed to inverted order of the NPs and PPs? Is this separately true for cases where NP-PP is the real ordering, and for cases where it is PP-NP? This will be tested by bootstrapping over language models and examples.
(2) Is the listeners memory-surprisal tradeoff significantly better when placing the shorter constituent (of this NP and PP) before the longer one?
Also tested by bootstrapping over language models and examples.

6) Describe exactly how outliers will be defined and handled, and your precise rule(s) for excluding observations.
Language models with overall validation average surprisal > 10 will be excluded (this didn't happen).

7) How many observations will be collected or what will determine sample size?
20 LSTM language models are constructed on a version of Penn Treebank where for all verb-initial VPs, the order of the remaining daughter constituents is shuffled.

Is the listener's memory-surprisal tradeoff significantly better on real as opposed to inverted order of the NPs and PPs?

Is this separately true for cases where NP-PP is the real ordering, and for cases where it is PP-NP?

Is the listeners memory-surprisal tradeoff significantly better when placing the shorter constituent (of this NP and PP) before the longer one?

%
%
%\section{NP Structure}
%
%- adjective order
%
%- noted by ...: Det - Num - Adj
%
%
%This is a case where the predictions differ from DepL
%
%
%E.g., if the choice of determiners, numerals, and adjectives are statistically independent when conditioning on the noun, will predict the effect.
%Computationally verify the effect.

@@

Ordering grammars are defined as in the Optimization paper: For each UD relation type, define a weight. To get random baseline languages, randomize the weights.
All grammars are valuated deterministically.

%Things that one may ponder: (1) what are even stronger baselines? (2) probably should do a control experiment using maximum-likelihood fits

%\subsection{Prediction on the level of POS}
%As mutual information cannot be realiably estimated on many of the smaller corpora in the UD project, we first compute (\ref{eq:memory-bound}) on the level of POS tags -- that is, we compute memory need for incremental prediction of POS sequences. %, following~\cite{futrell-noisy-context-2017}.
%As the summation in~(\ref{eq:memory-bound}) terminates when $t$ is greater than the length of any sentence, we can compute this quantity exactly for the empirical distribution over sentences in a given corpus.
%
%%We computed these quantitities for the optimized languages, the original languages, and for baseline  languages. For each language, we created 50 baseline ordering models, for which $a_\tau, b_\tau$ where drawn uniformly from $[-1, 1]$.
%%As mutual information between words is decreased if the language exhibits additional randomness (as our stochastic ordering models do) -- which decreases memory demands but increases entropy -- we hold the entropy of our languages constant by replacing each sampling step with taking the maximum -- that is, we take the zero temperature limit.
%%This ensures that all versions of a given language have the same entropy, and makes the numbers for memory demand comparable.
%%
%%%\mhahn{TODO results across languages}
%%
%%%We find that, averaged across all 50 languages, optimized models require less memory than 93 \% of the respective baseline models.
%
%Results for four languages are shown in Figure~\ref{fig:memory-results}.
%Averaged across languages, natural language requires less memory than 87 \% of random baseline ordering models.
%%Languages which require less memory than all baseline languages had significantly lower word order freedom ($t = -4.061$, $p = 0.00018$).
%
%
%%%For half of the languages
%%For 75 \% of all languages, the LSTM-optimized version takes less memory than 95 \% of the baseline models.
%%For 37 \% of the languages, it also requires less memory than the original language.
%
%%We found that, as predicted, memory usage in real languages was correlated with word order freedom:
%
%
%%\begin{figure*}
%%\includegraphics[width=0.98\textwidth]{figures/memory-demand-upos-facets.pdf}
%%%
%%\caption{Memory demands of real, optimized (for ease of LSTM language modeling), and baseline versions of languages computed using~(\ref{eq:memory-bound}): Optimizing for ease of language modeling results in languages that require less memory in incremental prediction. On the level of POS tags, languages with fixed word order minimize memory need, while languages with very free word order, such as Basque, have similar memory requirements to baseline languages.}\label{fig:memory-results}
%%\end{figure*}
%%

@@

This experiment is not fully confirmatory, since some data was collected for some languages before the predictions and the setup of the experiment were finalized.

@@

\paragraph{Speaker Memory: Distance from Pareto Frontier}
We want to test: Are languages closer to the Pareto frontier than baseline languages tend to be?

For this, we need a way of quantifying distance to the Pareto frontier.
We suggest to measure the distance as the fraction of baseline languages that are better on both dimensions.

More formally, fixing a natural language, let $P_0$ be the distribution of baseline languages mapped to the surprisal-memory plane, viewed as a probability distribution over the positive quadrant of the plane $\mathbb{R}_+^2$.
For any point $x \in \mathbb{R}_+^2$, let $A_x$ be the fraction of baseline languages that are better on both dimensions.
Similarly, let $B_x$ be the fraction of baseline languages that are worse on both dimensions.
Using the probability distribution $P_0$, we can write:
$$A_x := P_0(\{X : X_1 \leq x_1 \wedge X_2 \leq x_2\})$$
$$B_x := P_0(\{X : X_1 \geq x_1 \wedge X_2 \geq x_2\})$$
%Ways of using this to test and quantify for distance (I think they are all reasonable and worth reporting in SI. Question is which ones to rely on mostly for the main part):
Futher, let $P_1$ be the distribution in the surprisal-memory plane that is generated by the results obtained for the real orderings, with different random initializations.
For each language, we \emph{test} the hypothesis that the language optimizes the tradeoff by considering the null hypothesis that the real orderings do not Pareto-dominate more baseline orderings than there are baseline orderings dominating them:
$$\E_{x \sim P_1} [B_x] \leq \E_{x \sim P_1}[A_x]$$

We propose to \emph{quantify} closeness to the frontier using the following two statistics:

\begin{enumerate}
%	\item CIs for $C_x := B_x - A_x$ may be a resonable stopping criterion
	\item  $E := \frac{\E_{x \sim P_1}[A_x]}{\E_{x \sim P_1}[A_x+B_x]}$

		This has the following interpretation: Among orderings that either dominate or are dominated by the real orderings, what fraction is dominated by the real orerings?
		This quantity is $1$ if the real orderings are not Pareto-dominated by any baseline order, but do dominate some baseline orders.
		The null hypothesis defined in our test amounts to the inequality $\E_{x \sim P_1}E_x \leq 0.5$.

A drawback of this quantity is that it is meaningless when the real orders do not dominate, and are not dominated by, any real order.
The following quantity is always maningful:

	\item Compute $A_{x'}$ for each $x'$ from the baseline distribution $P_0$, and then calculate the percentile rank of $\E_{x \sim P_1}[A_x]$ among these values:
		$$D := P_0(A_X \leq \E_{x \sim P_1}[A_x])$$
This statistic describes how many random orders achieve closer distance to the frontier as measured by $A_x$.
		Stated differently, it quantifies how `extraordinary' the value $\E_{x \sim P_1}[A_x]$ is in relation to values typically attained for $x' \sim P_0$.
The statistic is close to zero if $\E_{x \sim P_1}[A_x]$ is smaller than for almost all samples from $P_0$.

We will obtain confidence intervals for the statistics via bootstrapping, and use a confidence interval for $B_x-A_x$ for testing $B_x \leq A_x$.
In the appendix, we report simulations that justify the use of bootstrapping in this situation.

@@

For each language, we collected at least 20 data points for real and baseline orderings each, and continued collecting until the CI for $D$ had width $\leq 0.3$, and the CI for $W$ had width $\leq 0.15$.
We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.

@@

Results are reported in Appendix~\ref{sec:results-table}.

\paragraph{Surprisal and Speaker Memory}
In all but X languages, the null hypothesis $\E_{x \sim P_1} [B_x] \leq \E_{x \sim P_1}[A_x]$ was rejected.
Also, in all but X languages, $D < 0.5$.

\paragraph{Surprisal and Listener Memory}
In all but X languages $\E[W] > 0.5$ significantly.

@@

@@

-- UD$\_$Polish-LFG (released in 2.2, not included in original experiment)

@@

@@

@@

-- another Vietnamese dependency treebank \citep{nguyen-bktreebank:-2017}

@@

-- furher (2.2/2.3 Japanese treebanks)

@@

\begin{proof}
	The number of bits of memory is given by the entropy $H[S_t]$.
	In view of (\ref{eq:markov-speaker}) and the Data Processing Inequality, this is lower bounded by the mutual information between past and future observations, $I[(X_i)_{i>0}, (X_i)_{i \leq 0}]$. 
Then the formula is a consequence of the chain rule for mutual information:
\begin{align*}
	I[(X_i)_{i > 0}, (X_i)_{i \leq 0}] &= \sum_{i=1}^\infty I[X_i, (X_i)_{i \leq 0}|X_1, ..., X_{i-1}] \\
	&= \sum_{i=1}^\infty \sum_{j=0}^{-\infty} I[X_i, X_j|X_{j+1}...X_{i-1}] \\
& = \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
\end{align*}
where we used stationarity of the process in the last step.
\end{proof}

@@

\section{Bootstrapping for Distance to Frontier}

TODO write up

We conducted simulations in the following setting:

sample two ellipses such that

(1) $\mu_1 \sim Unif([-0.5, 0.5]^2)$

(2) $\mu_2 \sim Unif([-1.5, 1.5]^2)$

$A_1, A_2 \sim Unif(\mathbb{R}^2 \times \mathbb{R}^2)$, $\Sigma_i = A_i A_i^T$ subject to the wo eigenvalues differing by a factor of $\leq 2$, and $\Sigma_1$ having area 1, $\Sigma_2$ having area 1/20.

Computed ground-truth $A_x, B_x$ from 3000 samples from both ellipses.

Bootstrapping: get 0.001, 0.999

Results:

For $A_x-B_x$: good coverage

For $E_x$: intervals are unreliable if $A_x + B_x$ is small

For $D_x$: good coverage

@@

\section{Thoughts}
Upper bound: Use $TH[X_t] - \sum_{t=1}^T (T-t) I_t = T (H[X_t] - \sum_{t=1}^T I_t) + \sum_{t=1}^T t I_t$ bits of memory and incur $\sum_{t>T} I_t$ in additioal surprisal.

\begin{tabular}{cccccccccccccc}
Afrikaans & Amharic & Arabic & Armenian & Basque & Breton & Bulgarian & Buryat \\
\includegraphics[width=0.14\textwidth]{figures/Afrikaans-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Amharic-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Arabic-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Armenian-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Basque-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Breton-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Bulgarian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Buryat-Adap-listener-surprisal-memory.pdf}
\\
Cantonese & Catalan & Chinese & Croatian & Czech & Danish & Dutch & Estonian \\
\includegraphics[width=0.14\textwidth]{figures/Cantonese-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Catalan-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Chinese-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Croatian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Czech-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Danish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Dutch-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Estonian-listener-surprisal-memory.pdf}
\\
Faroese & Finnish & French & German & Greek & Hebrew & Hindi & Hungarian \\
\includegraphics[width=0.14\textwidth]{figures/Faroese-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Finnish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/French-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/German-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Greek-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Hebrew-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Hindi-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Hungarian-listener-surprisal-memory.pdf}
\\
Indonesian & Japanese & Kazakh & Kurmanji & Latvian & Naija & North Sami & Norwegian \\
\includegraphics[width=0.14\textwidth]{figures/Indonesian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Japanese-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Kazakh-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Kurmanji-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Latvian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Naija-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/North_Sami-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Norwegian-listener-surprisal-memory.pdf}
\\
Persian & Polish & Romanian & Serbian & Slovak & Slovenian & Spanish & Swedish \\
\includegraphics[width=0.14\textwidth]{figures/Persian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Polish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Romanian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Serbian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Slovak-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Slovenian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Spanish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Swedish-listener-surprisal-memory.pdf}
\\
Tamil & Thai & Turkish & Ukrainian & Uyghur & Vietnamese \\
\includegraphics[width=0.14\textwidth]{figures/Tamil-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Thai-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Turkish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Ukrainian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Uyghur-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Vietnamese-listener-surprisal-memory.pdf}
\\\end{tabular}

@@

\appendix
\section{Bounding Memory in Imperfect Modeling}
We derived the nonlocality $M$ as a lower bound on the memory of a model that does optimally well in predicting the input process.
What happens when we consider imperfect (or near-optimal) prediction -- which may be a more realistic model for human processing?

The goal of this section is to show that $M$ is not just a lower bound on the memory required for perfect modeling, but also can be used to derive lower and upper bounds on imperfect modeling by suboptimal models.
To study imperfect modeling, we think of the process $(X_t)$ as fixed, and of a model as a function $f$ mapping histories $X_{<t}$ to memory representations, and computing a probability distribution over the vocabulary for $X_t$ based on the memory representation $f(X_{<t})$.
The memory representation $f(X_{<t})$ is a random variable whose entropy is the memory cost of the model.
In general, a model could use any mapping from memory states to distributions over the vocabulary to make its predictions.
However, we will focus on Bayes-optimal models, i.e., models that correctly compute the distribution of $X_t$ given the memory representation $f(X_{<t})$.
(Other models would incur even higher average surprisal, without reduction in memory cost.)
The average surprisal experienced by such a model is $H[X_t | f(X_{<t})]$, and is lower bounded by the true surprisal, $H[X_t|X_{<t}]$.

This section provides some results relating the model's memory cost $H[f(X_{<t})]$ to the attained surprisal $H[X_t | f(X_{<t})]$.
%Naturally, we expect that lower surprisal rates will require more memory.

%\subsection{Lower bound on approximation with n-th order Markov models}
%Let us assume that there is some $T$ such that $I_t = 0$ for $t > 0$ (e.g., $T$ is the length of the longest sentence in the corpus).
%We know $\sum_{t>T} I_{t}$ 
%If a Markov model has order $n$

There is a direct and obvious connection between the decay of mutual information and approximation by $n$-th order Markov models:
\begin{proposition}\label{prop:markov}
The Bayes-optimal $n$-th order Markov model will experience average surprisal (that is, cross-entropy) exactly $$H[X_t|X_{<t}] + \sum_{t > n} I[X_t, X_0 | X_{1\dots t-1}]$$
\end{proposition}

In the rest of this section, we think about further bounds for the case when we have information about memory, but not necessarily about the decay of mutual informations or the nature of the approximating model.

\subsection{Lower bound on imperfect approximation} 
Here we lower-bound the surprisal experienced by a model about which we only know that it has memory $J$ less than $M$, and where have no further knowledge about its structure.

As one might hope, models with memory $J$ less than $M$ will necessarily show greater average surprisal than a perfect model (simply because they cannot be perfect models by the previous proposition).
How much greater will their surprisal be? In general, if a model has memory much smaller than $M$, we cannot conclude that it must have very high average surprisal.\footnote{Example: consider, for each $T \in \mathbb{N}$, a process where $I_{T} = 1$, and $I_{t} = 0$ for $t \neq T$. We have $M = T$, but, even as $T \rightarrow \infty$, a model with zero memory achieves a KL loss of only 1. In this case, Proposition \ref{prop:lower-bound} yields the correct result.}
However, we can get such a bound in relation to the contribution of long-term dependencies to $M$.
%The state I've arrived at so far is still a bit complicated, but there is a reasonable final result for a wide class of processes.
%For smaller contribution of long-term dependencies is to $M$, we can lower bound the model's surprisal more strongly.
We write $$I_t := I[X_t, X_0 | X_{1\dots t-1}]$$

The following result holds for any process:
\begin{proposition}[Approximation by Models with Small Memory, lower bound]~\label{prop:lower-bound}
%For any $T \in \mathbb{N}$, a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \sum_{t \leq T} \frac{t}{T} I_t + \sum_{t > T} I_t - J/T$$
%Let $\epsilon > 0$, $T \in \mathbb{N}$ such that $\sum_{t > T} t I_t < \epsilon$.
%$$\sum_{t > T} t I[X_t, X_0 | X_{1\dots t-1}] < \epsilon$$
%Then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T} + \sum_{t > T} I_t$$
%Choose any $T \in \mathbb{N}$.
Any model with memory $J$ will experience average surprisal (that is, cross-entropy) at least
$$H[X_t|X_{<t}] +\sup_{T\in \mathbb{N}}\ \left[ \frac{\left(\sum_{t=1}^T t I_t\right)-J}{T} + \sum_{t > T} I_t\right]$$
%$$H[X_t|X_{<t}] + \sup_{T\in \mathbb{N}}\ \left[\frac{M-J}{T} - \sum_{t > T} (t/T - 1) I_t\right]$$
\end{proposition}
%The second term confirms the intuition that the model's surprisal should grow with the difference $M-J$, but the strength of the bound depends on the choice of $T$.\footnote{The bound can be vacuous for small choices of $T$: If we choose $T = 1$, the bound simply evaluates to the trivial lower bound $H[X_t] - J$, which may be even weaker than the likewise trivial lower bound $H[X_t|X_{<t}]$. It will become clear below that the bound is always nonvacuous when $T$ is large enough.}
%An equivalent way of stating this bound is:
Finding the optimal $T$ will require knowledge of the decay profile $I_t.$\footnote{For any fixed $T$, one can find a decay profile for which the bound would end up vacuous at that $T$, but meaningful for a larger $T$.}
Note that this bound only requires knowledge of $J$ and $I_t$, but no further knowledge of the process and the model.
I think that the bound is also optimal among all bounds that only require knowledge of $J$ and $I_t$, in the sense that for every choice of $I_t$ and $J$, there is a process and a model for which the bound is attained.\footnote{There can be no general nonvacuous bound relying only on $M$ and $J$, as shown in a previous footnote.}

\begin{proof}
Let $f$ be the function mapping histories $X_{<t}$ to memory representations of the model. By assumption, $f(X_{<t})$ can be encoded with up to $J$ bits, so $H[f(X_{<t})] \leq J$.
As explained above, without loss of generality, we can assume that the model is Bayes optimal. %: the model's distribution over $X_t$ after reading $X_{<t}$ is equal to $P(X_t | f(X_{<t}))$ (any other model would incur even higher surprisal).
By definition, the difference between the model's surprisal and optimal surprisal is
\footnote{The first inequality seems really crude. In general, we cannot do better. If we know more about the approximating model, we might be able to improve this. Also, consider the relation to the other inequality. Can they be attained at the same time?}
\begin{align*}
%& - \E_{X} \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t})\ \ \ \ \ \ \text{(for any $t$, due to stationarity)} \\
%& = - \frac{1}{T} \E_{X} \sum_{t=1}^T \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t}) \\
%& \geq  - \frac{1}{T} \E_{X} \log P(X_{1\dots T} | f(X_{<0})) - \log P(X_{1\dots T} | X_{<0})  \\
& H[X_t | f(X_{<t})] - H[X_t | X_{<t}] =  \frac{1}{T} \sum_{t=1}^{T} \left(H[X_t | f(X_{<t})] - H[X_t | X_{<t}]\right) \\
& \geq   \frac{1}{T} \left(H[X_{1\dots T} | f(X_{\leq 0})] - H[X_{1\dots T} | X_{\leq 0}]\right)  = \frac{1}{T} \left(I[X_{1\dots T}|X_{\leq 0}] - I[X_{1\dots T}|f(X_{\leq 0})]\right) \\
& = \frac{1}{T} \left(\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{\leq 0})]\right) = \frac{1}{T} \left(M - \sum_{t > T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{\leq 0})]\right) \\
& \geq \frac{1}{T} \left(M + \sum_{t > T} (T-t) I_t - J\right) 
\end{align*}
\end{proof}
%We can restate this bound in the following two, maybe more insightful, ways:\footnote{Note that (1) and (2) make clear that the bound is guaranteed to be larger than the basic bound $H[X_t|X_{<t}]$ when $T$ is chosen sufficiently large, but the gain over that trivial bound diminishes as $T \rightarrow \infty$. Finding an optimal $T$ that makes the bound as large as possible will depend on the process in question.}
%
%\begin{corollary}
%(1) A model with memory $J$ will experience average surprisal at least $$H[X_t|X_{<t}] +\sup_{T\in \mathbb{N}}\ \left[ \frac{\left(\sum_{t=1}^T t I_t\right)-J}{T} + \sum_{t > T} I_t\right]$$
%
%(2) Let $\epsilon > 0$, $T \in \mathbb{N}$ such that $\sum_{t > T} t I_t < \epsilon$.
%Then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T} + \sum_{t > T} I_t$$
%
%\end{corollary}
%
%

Here is way of phrasing the same statement in terms of memory:
\begin{corollary}
Let $\epsilon > 0$. Then it takes at least (and possibly more than) $$\sup_{T \in \mathbb{N}} \left(\sum_{t=1}^T t I_t\right) + T\left(\sum_{t > T} I_t - \epsilon\right)$$ bits of memory to achieve surprisal $\leq H[X_t|X_{<t}] + \epsilon$.
\end{corollary}

>>>>

\title{Word Orders Trade Off Predictability and Memory Load Across Languages}%  Word Order Minimizes  Across Languages
\author{Michael Hahn, Judith Degen, Richard Futrell}
\date{2018}

@@

\section{Memory-Surprisal Tradeoff}

With minimal assumptions, we will use information theory to derive a tradeoff between \emph{listener memory} and (listener) surprisal

In the second part of the paper, we examine whether word orders in natural language optimise this tradeoff.

@@

We now analyze memory from the perspective of the listener, who needs to maintain information about the past to predict the future.
The listener maintains a memory state $L_t$; there are no assumptions about the memory architecture and the nature of its computations.
%We assume that the listener has no access to the speaker's state beyond what the speaker has already uttered.
We only make two basic assumptions about the flow of information:

@@

@@

@@

\section{Relation to Theories of Sentence Processing}

@@

\paragraph{Statistical Complexity}
There are deep connections between our formalization of listener memory and studies of dynamic systems in the Physics literature.
Our formalization of listener memory, under the assumption of optimal prediction, is equivalent to the notion of \emph{Statistical Complexity} \citep{crutchfield-inferring-1989}.
Speaker memory corresponds to \emph{Generative Complexity} \cite{loehr-non-sufficient-2008, loehr-predictive-2010}.
The tradeoff between listener memory and surprisal is formally equivalent to the \emph{Recursive Information Bottleneck} considered by \cite{still-information-2014}.
The quantity in (\ref{eq:memory-bound}) is equal to the \emph{excess entropy}, and it has long been known to bound statistical and generative complexity \citep{crutchfield-inferring-1989}.
However, the link between memory and locality provided by our series expansion in (\ref{eq:memory-bound}) appears to be a novel contribution.

@@

@@

\paragraph{Ordering Grammrs}
We define ordering grammars, small interpretable models of dependency tree linearization.

Every dependency type $\tau$ is assigned a parameter $a_\tau \in [0,1]$.
Dependency trees are linearized by the following algorithm:

%Every dependency type $\tau$ is assigned a direction parameter $\alpha_\tau$ and a distance parameter $\beta_\tau$.

\paragraph{Baseline Languages}
We define baseline grammars by randomly sampling the parameters $a_\tau$.

@@

@@

@@

For each language, we collected at least 20 data points for real and baseline orderings each, and continued collecting until the CI for $W$ had width $\leq 0.15$.
We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.

@@

In all but X languages $\E[W] > 0.5$ significantly.

@@

\section{Experiment with MLE Grammars}

\paragraph{Maximum-Likelihood Grammars}
We additionally create ordering grammars that are fit to the actual orderings of each language.
These grammars faithfully represent the ordering rules if the actual language, to the extent that is possible in the formalism of ordering grammars.

We construct these grammars by interpreting ordering grammars probabilistically, and setting the parameters

\section{N-Gram Models}

\section{Morphology-Based Encoding}

@@

-- UD$\_$Polish-LFG (released in 2.2, not included in original experiment) (13,744 sentences)

@@

\section{Character-Level Modeling}

\section{Non-UD Dependency Treebanks}

@@

-- another Vietnamese dependency treebank \citep{nguyen-bktreebank:-2017} (5,639 sentences)

-- another Chinese dependency treebank LDC2012T05

\section{Constituency Treebank}

@@

@@

@@

@@

@@

@@
